Trion ARPES signature

### 2024-04-28 paper/main.tex
minor update
more formal figures
update julia env and code

\usepackage{textgreek}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{physics}
\usepackage{tensor}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{autobreak}
\usepackage[colorlinks,unicode]{hyperref} % , linkcolor=black, anchorcolor=black, citecolor=black, urlcolor=black, filecolor=black
\usepackage{xurl}
\usepackage[most]{tcolorbox}

@@

\title{Trion ARPES signature}
\author{Jinyuan Wu}
\email{jinyuan.wu@yale.edu}
\affiliation{Department of Mechanical Engineering and Materials Science, Yale University, New Haven, CT 06520}
\author{Diana Y. Qiu}
\email{diana.qiu@yale.edu}
\affiliation{Department of Mechanical Engineering and Materials Science, Yale University, New Haven, CT 06520}

@@

\section{Two-band model for trion}

In this work, we generalize the two-band model approach to trions \cite{chang2021variationally}
to the indirect band gap case.
We start with the following example.
\begin{equation}
    \begin{aligned}
        H &= \frac{(\ke - \vb{w})^2}{2\me} + \Eg 
        + \frac{\khi{1}^2}{2\mh} + \frac{\khi{2}^2}{2\mh} \\
        &\quad+ V(\rhi{1} - \rhi{2}) - V(\re - \rhi{1}) - V(\re - \rhi{2}).
    \end{aligned}
    \label{eq:two-band-w}
\end{equation}
Here $\khi{$i$}$ is the crystal momentum of the quantum state containing hole $i$ only, 
which is the opposite of the momentum of the corresponding electron state of the hole;
similarly, $\vb{k}^2/2\mh$ is the opposite of the dispersion relation of the valence band top,
and therefore $\mh > 0$, 
and under this Hamiltonian we treat holes as ordinary particles with positive masses and positive charges and ordinary momentum conservation relations without minus signs.

@@

We can define the canonical momenta of $\vb{r}_{1,2}$, 
hereafter referred to as $\vb{k}_{1,2}$, 
as $- \ii \hbar_{\vb{r}_{1, 2}}$;
in this way 
\begin{equation}
    \ke = \frac{\me}{M} \vb{P} - \vb{k}_1 - \vb{k}_2, \quad 
    \khi{1,2} = \vb{k}_{1, 2} + \frac{\mh}{M} \vb{P},
\end{equation}
and the kinetic energy part of the \eqref{eq:two-band-w} now reads 
\[
    H_{\text{kin}} = \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg 
    + \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mh} 
    - \frac{\vb{w}^2}{2M} + \frac{(\vb{k}_1 + \vb{k}_2 + \vb{w})^2}{2\me}.
\]
We however want to shift the definitions of $\vb{k}$ 
so that the kinetic part resembles the kinetic energy in a direct band gap system.
By redefining 
\begin{equation}
    \vb{k}_{1,2} = - \ii \hbar \partial_{\vb{r}_{1, 2}} + \frac{\vb{w}}{1 + \me / \mu}, \quad 
    \mu = \frac{\me \mh}{\me + \mh},
\end{equation}
and therefore 
\begin{equation}
    \begin{aligned}
        &\ke = \frac{\me}{M} \vb{P} + \frac{2 \mh \vb{w}}{M} - \vb{k}_1 - \vb{k}_2, \\ 
        &\khi{1,2} = \vb{k}_{1,2} - \frac{\mh \vb{w}}{M} + \frac{\mh}{M} \vb{P},
    \end{aligned}
\end{equation}
the total Hamiltonian now reads
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{2\me}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}}.
    \end{aligned}
    \label{eq:internal-ham}
\end{equation}
The redefinition of $\vb{k}_{1, 2}$ does not change the commutation relation 
between $\vb{r}_{1, 2}$ and $\vb{k}_{1, 2}$.
We find that the internal part of the Hamiltonian has the same form 
as the trion Hamiltonian in \cite{chang2021variationally}, 
which includes no indirect band gap.
This can also be seen as generalization of the indirect band gap exciton two-band model to the trion;
for the exciton case we have \cite{rustagi2018photoemission} 
\begin{equation}
    H = \frac{(\vb{Q} - \vb{w})^2}{2M} + \Eg + \frac{\vb{k}^2}{2\mu} - V(\vb{r}), 
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        &M = \me + \me, \\
        &\ke = \frac{\me}{M} \vb{Q} + \vb{k} + \frac{\mh}{M} \vb{w}, \\
        &\kh = \frac{\mh}{M} \vb{Q} - \vb{k} - \frac{\mh}{M} \vb{w}.
    \end{aligned}
\end{equation}

@@

@@

When $\vb{k}_1 = \vb{k}_2$, we have 
\begin{equation}
    \khi{1} = \frac{1}{2} (\vb{P} - \vb{w} - \ke), \quad 
    \khi{2} = \frac{1}{2} (\vb{P} + \vb{w} - \ke), 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = E_{S\vb{P}} + \Eg + \frac{(\vb{P} - \vb{w} - \ke)^2}{4\mh}.
\end{equation}

@@

The shifts of the center and the curvature of ARPES signatures are even more clearly 
when the trion mode has two electrons and one hole.
A typical setting of such a trion mode can be described by the following two-band model, 
where the two electrons reside at different valleys:
\begin{equation}
    \begin{aligned}
    H &= \frac{\kei{1}^2}{2\me} + \frac{(\kei{2} - \vb{w})^2}{2\me} + \frac{\kh^2}{2\mh} + 2 \Eg \\
    &\quad + V(\rei{1} - \rei{2}) - V(\rh - \rei{1}) - V(\rh - \rei{2}).
    \end{aligned}
\end{equation}
The decomposition of the Hamiltonian into internal and external terms 
follows exactly the same procedure as in the two-peak model; 
we only need to swap the positions of electrons and holes. 
Therefore we find that under 
\begin{equation}
    \begin{aligned}
        \vb{r}_1 &= \rei{1} - \rh, \quad 
        \vb{r}_2 = \rei{2} - \rh, \\
        \vb{k}_1 &= - \ii \partial_{\vb{r}_1} + \frac{\me}{M} \vb{w}, \quad 
        \vb{k}_2 = - \ii \partial_{\vb{r}_2} - \frac{\me + \mh}{M} \vb{w}, 
    \end{aligned}
\end{equation}
or conversely 
\begin{equation}
    \begin{aligned}
        &\kh = \frac{\mh}{M} (\vb{P} - \vb{w}) - \vb{k}_1 - \vb{k}_2, \\
        &\kei{1} = \vb{k}_1 - \frac{\me}{M} \vb{w} + \frac{\me}{M} \vb{P}, \\
        &\kei{2} = \vb{k}_2 + \frac{\me + \mh}{M} \vb{w} + \frac{\me}{M} \vb{P} ,
    \end{aligned}
\end{equation}
the Hamiltonian of the two-electron-one-hole trion is reduced to 
% See src/two-electron.nb
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + 2 \Eg, \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2}{2\mu} + \frac{\vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{2\mh}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}},
    \end{aligned}
\end{equation}
a form resembling \eqref{eq:internal-ham}, 
the only differences being the coefficient of the $\vb{k}_1 \cdot \vb{k}_2$ term
and the $2\Eg$ term due to the fact that there are two electrons now.
Now when one electron is driven out by the probe, 
it can be e1 or e2, so $\kp = \kei{1,2}$, 
momentum conservation blocks any block terms between the two valleys, 
so we can evaluate the ARPES signature contributed by the two valleys separately.
This drastically changes the dispersion relation of the ARPES signature, 
but the structure of trion wave function is intact.
The most remarkable difference between the ARPES signature of a negative trion 
and that of a positive trion
is the former has a vertical ``tail'': 
when $\kp = \kei{1}$ is chosen, $\vb{k}_2$ is allow to move freely, 
and it continuously adjusts the energy of the remaining of the trion after e1 is driven out, 
which means the ARPES signature will have a long, vertical component 
that does not coincide with any dispersion relations. % TODO: rework the wording

Therefore, the curvature of the ARPES signature can still be found by either postulating $\vb{k}_1 = \vb{k}_2$ or by postulating that one of the momenta vanishes.
When $\vb{k}_1 = \vb{k}_2$ and $\kp = \kei{1}$, we have 
\begin{equation}
    \kei{2} - \vb{w} = \kp, \quad \kh = \vb{P} - \vb{w} - 2 \kp, 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = \Eg + E_{\text{b}} + \frac{\vb{P}^2}{2M} - \frac{\kp^2}{2\me} - \frac{(2 \kp + \vb{w} - \vb{P})^2}{2\mh}, 
\end{equation}
while the effective mass is

@@

\bibliography{arpes,trion}

>>>>

\usepackage{textgreek}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{physics}
\usepackage{tensor}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{autobreak}
\usepackage[colorlinks,unicode]{hyperref} % , linkcolor=black, anchorcolor=black, citecolor=black, urlcolor=black, filecolor=black
\usepackage{xurl}
\usepackage[most]{tcolorbox}

@@

\title{Trion ARPES signature}
\author{Jinyuan Wu}
\affiliation{Department of Mechanical Engineering and Materials Science, Yale University, New Haven, CT 06520}
\author{Diana Y. Qiu}
\email{diana.qiu@yale.edu}
\affiliation{Department of Mechanical Engineering and Materials Science, Yale University, New Haven, CT 06520}

@@

\section{Two-band model for a positive trion}

In this work, we generalize the two-band model approach to trions \cite{chang2021variationally}
to the indirect band gap case.
We start with the following Hamiltonian of a negative trion, 
i.e. a trion that contains one electron and two holes (ehh).
The effective Hamiltonian reads
\begin{equation}
    \begin{aligned}
        H &= \frac{(\ke - \vb{w})^2}{2\me} + \Eg 
        + \frac{\khi{1}^2}{2\mh} + \frac{\khi{2}^2}{2\mh} \\
        &\quad+ V(\rhi{1} - \rhi{2}) - V(\re - \rhi{1}) - V(\re - \rhi{2}).
    \end{aligned}
    \label{eq:two-band-w}
\end{equation}
Here $\khi{$i$}$ is the crystal momentum of the quantum state containing hole $i$ only, 
which is the opposite of the momentum of the corresponding electron state of the hole;
similarly, $\vb{k}^2/2\mh$ is the opposite of the dispersion relation of the valence band top,
and therefore $\mh > 0$, 
and under this Hamiltonian we treat holes as ordinary particles with positive masses and positive charges and ordinary momentum conservation relations without minus signs.

@@

We can define the canonical momenta of $\vb{r}_{1,2}$, 
hereafter referred to as $\vb{k}_{1,2}$, 
as $- \ii \hbar_{\vb{r}_{1, 2}}$;
in this way 
\begin{equation}
    \ke = \frac{\me}{M} \vb{P} - \vb{k}_1 - \vb{k}_2, \quad 
    \khi{1,2} = \vb{k}_{1, 2} + \frac{\mh}{M} \vb{P},
\end{equation}
and the kinetic energy part of the \eqref{eq:two-band-w} now reads 
\[
    H_{\text{kin}} = \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg 
    + \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mh} 
    - \frac{\vb{w}^2}{2M} + \frac{(\vb{k}_1 + \vb{k}_2 + \vb{w})^2}{2\me}.
\]
We however want to shift the definitions of $\vb{k}$ 
so that the kinetic part resembles the kinetic energy in a direct band gap system.
By redefining 
\begin{equation}
    \vb{k}_{1,2} = - \ii \hbar \partial_{\vb{r}_{1, 2}} + \frac{\vb{w}}{1 + \me / \mu}, \quad 
    \mu = \frac{\me \mh}{\me + \mh},
\end{equation}
and therefore 
\begin{equation}
    \begin{aligned}
        &\ke = \frac{\me}{M} \vb{P} + \frac{2 \mh \vb{w}}{M} - \vb{k}_1 - \vb{k}_2, \\ 
        &\khi{1,2} = \vb{k}_{1,2} - \frac{\mh \vb{w}}{M} + \frac{\mh}{M} \vb{P},
    \end{aligned}
\end{equation}
the total Hamiltonian now reads
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{\me}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}}.
    \end{aligned}
    \label{eq:internal-ham}
\end{equation}
The redefinition of $\vb{k}_{1, 2}$ does not change the commutation relation 
between $\vb{r}_{1, 2}$ and $\vb{k}_{1, 2}$.
We find that the internal part of the Hamiltonian has the same form 
as the trion Hamiltonian in \cite{chang2021variationally}, 
which includes no indirect band gap.
This can also be seen as generalization of the indirect band gap exciton two-band model to the trion;
for the exciton case we have \cite{rustagi2018photoemission} 
\begin{equation}
    H = \frac{(\vb{Q} - \vb{w})^2}{2M} + \Eg + \frac{\vb{k}^2}{2\mu} - V(\vb{r}), 
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        &M = \me + \me, \\
        &\ke = \frac{\me}{M} \vb{Q} + \vb{k} + \frac{\mh}{M} \vb{w}, \\
        &\kh = \frac{\mh}{M} \vb{Q} - \vb{k} - \frac{\mh}{M} \vb{w}.
    \end{aligned}
\end{equation}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/exciton-Q-0.2-w-0.0-makie.png}      
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/trion-e1h1h2-P-1.2-w-1.0-makie.png}
        \subcaption{}
    \end{subfigure}
    \caption{The simulated ARPES spectra of the exciton (a) and the ehh trion (b). 
    Effective masses and the band gap are from Ref. \cite{qiu2013optical}.
    The curvatures of the two signals are different, 
    and so are the center of them.}
\end{figure*}

@@

When $\vb{k}_1 = \vb{k}_2$, we have 
\begin{equation}
    \khi{1} = \frac{1}{2} (\vb{P} - \vb{w} - \ke), \quad 
    \khi{2} = \frac{1}{2} (\vb{P} + \vb{w} - \ke), 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = E_{S\vb{P}} + \Eg + \frac{(\vb{P} - \vb{w} - \ke)^2}{4\mh}.
\end{equation}
We note that the effective mass of this dispersion relation is $2 \mh$ instead of $\mh$,
meaning that the ARPES signature does not have a well-defined curvature corresponding to an effective mass of $\mh$.
Moreover, the center of the ARPES signature is predicted to be 
\begin{equation}
    \ke = \frac{\me}{M} (\vb{P} - \vb{w}),
\end{equation}
different from

@@

The shifts of the center and the curvature of ARPES signatures are even more clearly 
when the trion mode has two electrons and one hole.
A typical setting of such a trion mode can be described by the following two-band model, 
where the two electrons reside at different valleys:
\begin{equation}
    \begin{aligned}
    H &= \frac{\kei{1}^2}{2\me} + \frac{(\kei{2} - \vb{w})^2}{2\me} + \frac{\kh^2}{2\mh} + 2 \Eg \\
    &\quad + V(\rei{1} - \rei{2}) - V(\rh - \rei{1}) - V(\rh - \rei{2}).
    \end{aligned}
\end{equation}
The decomposition of the Hamiltonian into internal and external terms 
follows exactly the same procedure as in the two-peak model; 
we only need to swap the positions of electrons and holes. 
Therefore we find that under 
\begin{equation}
    \begin{aligned}
        \vb{r}_1 &= \rei{1} - \rh, \quad 
        \vb{r}_2 = \rei{2} - \rh, \\
        \vb{k}_1 &= - \ii \partial_{\vb{r}_1} + \frac{\me}{M} \vb{w}, \quad 
        \vb{k}_2 = - \ii \partial_{\vb{r}_2} - \frac{\me + \mh}{M} \vb{w}, 
    \end{aligned}
\end{equation}
or conversely 
\begin{equation}
    \begin{aligned}
        &\kh = \frac{\mh}{M} (\vb{P} - \vb{w}) - \vb{k}_1 - \vb{k}_2, \\
        &\kei{1} = \vb{k}_1 - \frac{\me}{M} \vb{w} + \frac{\me}{M} \vb{P}, \\
        &\kei{2} = \vb{k}_2 + \frac{\me + \mh}{M} \vb{w} + \frac{\me}{M} \vb{P} ,
    \end{aligned}
\end{equation}
the Hamiltonian of the two-electron-one-hole trion is reduced to 
% See src/two-electron.nb
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + 2 \Eg, \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2}{2\mu} + \frac{\vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{\mh}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}},
    \end{aligned}
\end{equation}
a form resembling \eqref{eq:internal-ham}, 
the only differences being the coefficient of the $\vb{k}_1 \cdot \vb{k}_2$ term
and the $2\Eg$ term due to the fact that there are two electrons now.
Now when one electron is driven out by the probe, 
it can be e1 or e2, so $\kp = \kei{1,2}$, 
momentum conservation blocks any block terms between the two valleys, 
so we can evaluate the ARPES signature contributed by the two valleys separately.
This drastically changes the dispersion relation of the ARPES signature, 
but the structure of trion wave function is intact.
The most remarkable difference between the ARPES signature of a negative trion 
and that of a positive trion
is the former has a vertical ``tail'': 
when $\kp = \kei{1}$ is chosen, $\vb{k}_2$ is allow to move freely, 
and it continuously adjusts the energy of the remaining of the trion after e1 is driven out, 
which means the ARPES signature will have a long, vertical component 
that does not coincide with any dispersion relations. % TODO: rework the wording

Therefore, the curvature of the ARPES signature can still be found by either postulating $\vb{k}_1 = \vb{k}_2$ or by postulating that one of the momenta vanishes.
When $\vb{k}_1 = \vb{k}_2$ and $\kp = \kei{1}$, we have 
\begin{equation}
    \kei{2} - \vb{w} = \kp, \quad \kh = \vb{P} - \vb{w} - 2 \kp, 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = \Eg + E_{\text{b}} + \frac{\vb{P}^2}{2M} - \frac{\kp^2}{2\me} - \frac{(2 \kp + \vb{w} - \vb{P})^2}{2\mh}, 
\end{equation}
while the effective mass is  TODO: an analytic expression giving the shape of the signature is still elusive

\begin{figure}
    \includegraphics[width=0.4\textwidth]{figs/trion-e1e2h1-P-1.0-w-1.0-makie.png}
    \caption{The simulated ARPES spectrum of the eeh trion. Effective masses and the band gap are from Ref.~\cite{qiu2013optical}.}
\end{figure}

@@

\bibliography{arpes,trion,materials}
